Kindly Bent to Free Us

### 2020-02-28 kindly/motivation-persistent.tex
motivation cosmetics

\cref{sig:hybarray} contains an API @HYBARRAY@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HYBARRAY@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
modifications that correspond to choice points using @set@,
which makes it trivial to come back to the previous version
of the array, while other modifications use @set_mut@, which
forbids uses of the previous version.

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the @Matrix@ type uses the same API as @PersistArray@
but with two indices.
Each cell contains an integer set that represent admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the @propagate@ function.
The @propagate@ function uses direct mutation through a
exclusive borrow of the matrix as it need not preserve the previous
version of the board.
The implementation of @propagate@ is split into three parts
for lines, columns, and square, which are all very similar to function
@propagate_lines@ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state @new_g@ created for one
particular branch of the choice.
As the @set@ function only requires an unrestricted shared borrow,
the closure @try_solution@ remains unrestricted even though
it captures the borrow @&g@.
The price is that @try_solution@ cannot escape from
@&g@'s region. In this example, inference places the region
around the @begin@/@end@ scope.
Hence, @try_solution@ can be used in
the iteration in \cref{line:iter}.
% As @&g@ is unrestricted, we could even use a parallel iteration
% instead of a sequential one.
As @g@ is linear we must free it outside of the region
before returning (\cref{line:free:g}).

>>>>

\cref{sig:hybarray} contains an API @HYBARRAY@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HYBARRAY@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function also
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

Our implementation of a Sudoku solver (\cref{ex:sudoku}) performs
modifications that correspond to choice points using @set@,
which makes it trivial to come back to the previous version
of the array, while other modifications use @set_mut@, which
cannot be reverted.

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the @Matrix@ type uses the same API as @PersistArray@
but with two indices.
Each cell contains an integer set that represents admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the @propagate@ function.
The @propagate@ function uses direct mutation through an
exclusive borrow of the matrix as it need not preserve the previous
version of the board.
The implementation of @propagate@ is split into three parts
for lines, columns, and square, which are all very similar to function
@propagate_lines@ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state @new_g@ created for one
particular branch of the choice.
As the @set@ function only requires an unrestricted shared borrow,
the closure @try_solution@ remains unrestricted even though
it captures the borrow @&g@.
The price is that @try_solution@ cannot escape from
@&g@'s region. In this example, the inferred region corresponds to the
@begin@/@end@ scope. 
Hence, @try_solution@ can be used in
the iteration in \cref{line:iter}.
% As @&g@ is unrestricted, we could even use a parallel iteration
% instead of a sequential one.
As @g@ is linear we must free it outside of the region
before returning (\cref{line:free:g}).

### 2020-02-28 kindly/motivation.tex
Move back the sudoku example.
motivation cosmetics

\lang{} is amenable to a functional resource-passing style
common in functional encodings of session types (e.g.,
\cite{DBLP:journals/jfp/Padovani17}; see also
\cref{sec:session-linearity} in the supplement). But it really shines
when manipulating mutable resources like buffers or connection pools
using a mixed functional and imperative programming style.
%
To support this usage pattern of linearity, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
we demonstrate first through the example of linear arrays to demonstrate
simple borrowing and imperative programming and second with iterators
on linear values to demonstrate advanced uses of regions.
Further examples are also available in \cref{sec:extra-example}.

@@

The API for mutable linear arrays (shown in \cref{sig:array})
aims to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
One would first use @create n v@ to create
an (unrestricted) array of size @n@ initialized with value
@v@. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the @map@ function we
can transform the unrestricted elements into linear (affine)
ones. This array can be processed further using @map@ (or @set@ if
affine). The @get@ function is only applicable if the element type
is unrestricted as one element is duplicated. The @length@
function is always applicable. To @free@ an array the elements must be
affine.

To manage the different kinds of accessing the array we distinguish between constructors, destructors,
observers, and mutators.
Constructors and destructors like @create@ and @free@ manipulate the whole
array. %As with channels in \cref{sec:session-linearity},
The constructor
@create@ yields a linear resource which is consumed by @free@.
During the lifetime of the array resource @a@, we can split
off \emph{shared borrows} @&a@ that provide a read-only
view or \emph{exclusive borrows} @&!a@ for read-write views.
Observer functions such as @length@ and @get@ expect an shared borrow
argument, mutator functions such a @set@ expect
an exclusive borrow.

@@

We can use the API 
to create an
array of Fibonacci numbers following an imperative coding style:
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} contains two shared borrows
in the same expression which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set} which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.
%
More precisely, here is is an annotated excerpt with explicit regions
marked in braces @{| ... |}@:
\begin{lstlisting}[firstnumber=3]
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}

A region consists of the header expression of the @let@ in
\cref{ex:array:region:get}. It is contained in another region
spanning the body of the @for@ loop. \lang guarantees that borrows
never escape the smallest enclosing region. It employs a system of
\emph{indexed kinds} like @aff_r@ and @un_r@ where
$r$ is a positive integer that corresponds to the lexical nesting
depth of regions. For instance, the type of @&!a@ in
\cref{ex:array:set} has kind @aff_1@ whereas the type of
@&a@ in \cref{ex:array:region:get} has kind
@un_2@ and the typing of the inner region is such that types with
kind indexes greater than or equal to $2$ cannot escape.
In the example, borrows cannot escape  because they are consumed
immediately by @get@ and @set@.

@@

We can then easily use this function to fold over an array of files. Here
we compute the sum of their sizes:
\begin{lstlisting}
let total_size = fold (fun f s -> File.size f + s) &all_files 0
\end{lstlisting}

Naturally, this is not sufficient if we want to mutate the array during iteration.
For this purpose, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}
val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
\end{lstlisting}

Since the callback will be called multiple times, it is explicitly
annotated with @un@.

While the distinction between mutable and immutable iteration functions
might seem unfortunate, it is typical of
programming with borrows and is also present in the Rust standard library.
It allows the programmer to explicitly state how various iterations
are allowed to compose and be optimized.
\lang's region variables ensures that the content iterated on can never
be leaked outside of the iteration function.
This pattern is essential in many use-cases of linearity such
as pools of linear objects (see \cref{tuto:pool}).

>>>>

\lang{} is amenable to a functional resource-passing style
common in functional encodings of session types (e.g.,
\cite{DBLP:journals/jfp/Padovani17}; see also
\cref{sec:session-linearity} in the supplement) as well as other
functional resource handling. But it really shines
when manipulating mutable resources like buffers or connection pools
using a mix of functional and imperative programming styles.
%
To support this usage pattern of linearity, \lang{} relies on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}. Our first
example of linear arrays demonstrates simple borrowing and
imperative programming; the second example is a Sudoko solver using a
hybrid copy-on-write data structure; the third example demonstrates advanced uses
of regions with iterators on linear values.
Further examples are available in \cref{sec:extra-example}.

@@

The API for mutable linear arrays (\cref{sig:array})
aims to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
A program would first use @create n v@ to create
an array of size @n@ initialized with value
@v@. The value @v@ must be unrestricted as it is duplicated to
initialize all array elements. Using the @map@ function we
can transform the unrestricted elements into linear (affine)
ones. The resulting array can be processed further using @map@ (or @set@ if
affine). The @get@ function is only applicable if the element type
is unrestricted as one element is duplicated. The @length@
function is always applicable. To @free@ an array the elements must be
affine.

To manage the different kinds of accessing the array we distinguish between constructors, destructors,
observers, and mutators.
Constructors and destructors like @create@ and @free@ manipulate the whole
array. %As with channels in \cref{sec:session-linearity},
The constructor
@create@ yields a linear resource which is consumed by @free@.
During the lifetime of the array resource @a@, we can split
off \emph{shared borrows} @&a@ that provide a read-only
view or \emph{exclusive borrows} @&!a@ for read-write views.
Observer functions such as @length@ and @get@ expect a shared borrow
argument, mutator functions such a @set@ expect
an exclusive borrow.

@@

Using the API 
we can create an
array of Fibonacci numbers in an imperative coding style:
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} contains two shared borrows
in the same expression which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set} which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.
%
More precisely, here is is an annotated excerpt with regions explicitly
marked by braces @{| ... |}@:
\begin{lstlisting}[firstnumber=3]
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}

One region consists of the header expression of the @let@ in
\cref{ex:array:region:get}. It is contained in another region
spanning the body of the @for@ loop. \lang guarantees that borrows
never escape the smallest enclosing region. It employs a system of
\emph{indexed kinds} like @aff_r@ and @un_r@ where
$r$ is a positive integer that corresponds to the lexical nesting
depth of regions. For instance, the type of @&!a@ in
\cref{ex:array:set} has kind @aff_1@ whereas the type of
@&a@ in \cref{ex:array:region:get} has kind
@un_2@ and the typing of the inner region is such that types with
kind indexes greater than or equal to $2$ cannot escape.
In the example, borrows cannot escape  because they are consumed
immediately by @get@ and @set@.

\input{motivation-persistent}

@@

As an easy example, we fold over an array of files to compute the sum of their sizes:
\begin{lstlisting}
let total_size = fold (fun f s -> File.size f + s) &all_files 0
\end{lstlisting}

This approach is not sufficient if we want to mutate the array during iteration.
To this end, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}
val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
\end{lstlisting}

The callback is explicitly annotated with @un@ because  it will be called multiple times.

While the distinction between mutable and immutable iteration functions
seems unfortunate, it is typical of
programming with borrows and is also present in the Rust standard library.
It enables the programmer to explicitly state how different iterations
may be composed and optimized.
\lang's region variables ensure that the content iterated on can never
be leaked outside of the iteration function.
This pattern is essential in many use cases of linearity such
as pools of linear objects (see \cref{tuto:pool}).
